---
title: "Cox mixed-effects models for genome-wide association analysis"
author: "Liang He"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message=FALSE, warning=FALSE,echo=FALSE}
library(knitcitations)
cleanbib()
options("citation_format" = "pandoc")
r<-citep("10.3389/fpubh.2016.00003")
write.bibtex(file="references.bib")
```


## Overview

Time-to-event is one of the most important phenotypes in genetic epidemiology. The R-package, "coxmeg", provides a set of utilities to fit a Cox mixed-effects model and to efficiently perform genome-wide association analysis of time-to-event phenotypes using a Cox mixed-effects model. 

## Installation

### Most-recent version
```{r,eval=FALSE}
install.packages("coxmeg", repos="http://R-Forge.R-project.org")
```

## Fit a Cox mixed-effects model with a sparse relatedness matrix

We illustrate how to use coxmeg to fit a Cox mixed-effects model with a sparse relatedness matrix. We first simulate a block-diagonal relatedness matrix for a cohort consisting of 200 families, each of which has five members.

```{r,echo=TRUE}
library(coxmeg)
library(MASS)
library(Matrix)
n_f <- 200
mat_list <- list()
size <- rep(5,n_f)
offd <- 0.5
for(i in 1:n_f)
{
  mat_list[[i]] <- matrix(offd,size[i],size[i])
  diag(mat_list[[i]]) <- 1
}
sigma <- as.matrix(bdiag(mat_list))
sigma = as(sigma,'dgCMatrix')

```

We use 'dgCMatrix' to save memory. Next, we simulate random effects and time-to-event outcomes assuming a constant baseline hazard function. We assume that the variance component is 0.2. We also simulate a risk factor with log(HR)=0.1. 

```{r,echo=TRUE}
n = nrow(sigma)
tau_var <- 0.2
x <- mvrnorm(1, rep(0,n), tau_var*sigma)
pred = rnorm(n,0,1)
myrates <- exp(x+0.1*pred-1)
y <- rexp(n, rate = myrates)
cen <- rexp(n, rate = 0.02 )
ycen <- pmin(y, cen)
outcome <- cbind(ycen,as.numeric(y <= cen))
head(outcome)
sigma[1:5,1:5]

```

We fit a Cox mixed-effects model using the ```coxmeg``` function. We set ```dense=FALSE``` to indicate that the relatedness matrix is sparse. However, the function will automatically treat it as dense if there are more than 50% non-zero elements in the matrix. We set ```order=1``` to use the first-order approximation of the inverse Hessian matrix in the optimization.  

```{r,echo=TRUE}
re = coxmeg(outcome,sigma,pred,order=1,dense=FALSE)
re
```

In the result, ```coxmeg``` is estimated variance component. We compare the results with coxme. The integrated log-likelihoods cannot be compared directly because different approximation of log-determinant is used.   

```{r,echo=TRUE}
library(coxme)
bls <- c(1)
for(i in (size[1]-1):1)
{bls <- c(bls, c(rep(offd,i),1))}
tmat <- bdsmatrix(blocksize=size, blocks=rep(bls,n_f),dimnames=list(as.character(1:n),as.character(1:n)))
re_coxme = coxme(Surv(outcome[,1],outcome[,2])~as.matrix(pred)+(1|as.character(1:n)), varlist=list(tmat),ties='breslow')
re_coxme
```


